tstrait: a quantitative trait simulator for ancestral recombination graphs

### 2023-11-28 paper.tex
Edit pass. Shuffle some intro content

Genome-wide association studies (GWAS)
identify genetic variants that are statistically
associated with a specific trait~\citep{uffelmann2021}.
Many loci that are associated with various human diseases and
traits have been
identified~\citep[e.g.][]{yengo2022,mathieson2023},
% \citep{locke2015,ishigaki2022,
% mahajan2022,yengo2022,mathieson2023},
and GWAS results are actively being incorporated into clinical
practice~\citep{visscher2017}.
This success has led to the collection of many population-scale Biobank
datasets~\citep{tanjo2021practical},
but such scale is presenting significant challenges to GWAS
methodology~\citep{uffelmann2021}.

Ancestral recombination graphs (ARGs) encode the interwoven paths
of genetic inheritance arising from
% Explanation of citations: people might get offended if we don't cite
% Griffiths when first mentioning the ARG, but Hudson has just as
% much to do with the actual definition that we use in Wong 2023.
recombination~\citep{hudson1983properties,griffiths1997ancestral,wong2023general},
and recent
breakthroughs in inference have opened many new possibilities for their
practical application~\citep{lewanski2023era}.
Building on these recent developments, substantial progress has been made
on statistical genetics applications.
For example,  ARG-based methods
can detect more ultra rare variants than conventional association testing
methods~\citep{zhang2023};
have better power to detect causal loci in
quantitative-trait locus mapping~\citep{link2023tree};
and can provide a sparse and efficient model of linkage disequilibrium
in GWAS and downstream applications~\citep{nowbandegani2023extremely}.

@@

An important property of these modern simulation methods is that
they output ARGs rather than genotypes \citep[e.g.\ as VCF,][]{danecek2011}.
Using the succinct tree sequence ARG encoding~\citep{wong2023general}
and the open-source \texttt{tskit} library~\citep{ralph2020},
these ARG methods can simulate genetic variation for millions of
samples and store the output very compactly.
For example, a highly realistic simulation
of chromosome 9 for 1.4 million
French-Canadian individuals~\citep{anderson2023}
requires around 550GB of storage space in gzip-compressed VCF.
At 1.36GB the original simulated ARG
(compressed using the \texttt{tszip} utility) is around 400X smaller.
Furthermore, many computations can
be expressed efficiently in terms of the underlying
ARG~\citep{kelleher2016efficient,ralph2020}, without
needing to decode the actual variation data.
Finally, outputting a simulated ARG provides access to the full
history, not just the genetic variation among the samples.

Although there are sophisticated methods available for simulating
ARGs at the population scale, there is currently no
easy way to simulate quantitative traits based on such an ARG.
Many existing methods to simulate quantitative traits from a given set of
genetic sequences assume that the genotypes fit in
memory~\citep[e.g.][]{meyer2018,fernandes2020}, which makes them
impractical at Biobank scale (e.g., the French-Canadian dataset discussed
above would require 140TB of RAM assuming 1 byte per genotype).
Methods that read parts of the genotype matrix from file as
required~\citep[e.g.][]{wharrie2023hapnest} can be used on
reasonable hardware, but working with such large files is slow
and cumbersome.
More fundamentally, exporting genotypes discards much of rich
detail about ancestral history contained in an ARG, and
it is exactly this information that we wish to take
advantage of when using inferred ARGs in GWAS applications.
In their analysis of the portability of polygenic risk
scores across populations, \citet{martin2017}
demonstrated the utility of simulating phenotypes directly
from an ARG. Their approach, however, is tightly coupled to the details
of the study and not designed to be reused.
Simulation code can be subtle and difficult to
debug~\citep{ragsdale2020lessons}, and there
is a critical need for a well-documented and
thoroughly tested means of simulating quantitative
traits directly from an ARG.

In this paper we present \texttt{tstrait}, a new Python library that
efficiently simulates quantitative traits on an arbitrary ARG
in \texttt{tskit} format.
\texttt{Tstrait} can
quickly simulate quantitative traits for population-scale datasets,
with a very low memory overhead. This approach also provides a
richer model than existing methods,
allowing us to, for example, simulate traits for
ancestral individuals as well as contemporary samples.
The \texttt{tstrait} library also integrates well with the wider
Python data-science ecosystem~\citep{numpy},
allowing users to efficiently analyse
large-scale data using familiar and ergonomic tools.

>>>>

Genome-wide association studies (GWAS)
identify genetic variants that are statistically
associated with a specific trait~\citep{uffelmann2021}.
Many loci that are associated with various human diseases and
traits have been
identified~\citep[e.g.][]{yengo2022,mathieson2023},
% \citep{locke2015,ishigaki2022,
% mahajan2022,yengo2022,mathieson2023},
and GWAS results are actively being incorporated into clinical
practice~\citep{visscher2017}.
The great success of GWAS has prompted the collection of many biobank
datasets consisting of hundreds of thousands of
participants~\citep{tanjo2021practical},
but this scale presents significant challenges to
current GWAS methodology~\citep{uffelmann2021}.

@@

An important property of these population genetic simulation methods is that
they output ancestral recombination graphs (ARGs)
rather than sample genotypes.
ARGs encode the interwoven paths
of genetic inheritance arising from
% Explanation of citations: people might get offended if we don't cite
% Griffiths when first mentioning the ARG, but Hudson has just as
% much to do with the actual definition that we use in Wong 2023.
recombination~\citep{hudson1983properties,griffiths1997ancestral,wong2023general},
and contain rich detail about ancestral processes.
Recent breakthroughs in inferrence methods have made it possible
to estimate ARGs at biobank scale~\citep{kelleher2019,zhang2023},
and there is now intense interest in their practical
application~\citep{lewanski2023era}.
Statistical genetics has been a particular focus,
and ARG-based methods have been shown
to detect more ultra rare variants than conventional association testing
methods~\citep{zhang2023};
have better power to detect causal loci in
quantitative-trait locus mapping~\citep{link2023tree};
and to provide a sparse and efficient model of linkage disequilibrium
in GWAS and downstream applications~\citep{nowbandegani2023extremely}.

ARG-based methods can simulate genetic variation for millions of
samples and store the output very compactly
using the ``succinct tree sequence'' encoding~\citep{wong2023general}.
and \texttt{tskit} library~\citep{ralph2020}.
For example, a highly realistic simulation
of chromosome 9 for 1.4 million
French-Canadian individuals~\citep{anderson2023}
requires around 550GB of storage space in gzip-compressed
VCF~\citep{danecek2011}.
At 1.36GB the original simulated ARG
(compressed using the \texttt{tszip} utility) is around 400X smaller.
Furthermore, many computations can
be expressed efficiently in terms of the underlying
ARG~\citep{kelleher2016efficient,ralph2020}, without
needing to decode the actual variation data.
Finally, outputting a simulated ARG provides access to the full
history, not just the genetic variation among the samples.

Although there are sophisticated methods available for simulating
ARGs at biobank scale, there is currently no
easy way to simulate quantitative traits based on such an ARG.
Many existing methods to simulate quantitative traits from a given set of
genetic sequences assume that the genotypes fit in
memory~\citep[e.g.][]{meyer2018,fernandes2020}, which makes them
impractical at biobank scale (the French-Canadian dataset discussed
above would require 140TB of RAM assuming 1 byte per genotype).
Methods that read parts of the genotype matrix from file as
required~\citep[e.g.][]{wharrie2023hapnest} can be used on
reasonable hardware, but working with such large files is slow
and cumbersome.
More fundamentally, exporting genotypes discards much of rich
detail about ancestral history contained in an ARG, and
it is exactly this information that we wish to take
advantage of when using inferred ARGs in GWAS applications.
In their analysis of the portability of polygenic risk
scores across populations, \citet{martin2017}
demonstrated the utility of simulating phenotypes directly
from an ARG. Their approach, however, is tightly coupled to the details
of the study and not designed to be reused.
Simulation code can be subtle and difficult to
debug~\citep{ragsdale2020lessons}, and there
is a critical need for a well-documented and
thoroughly tested means of simulating quantitative
traits directly from an ARG.

In this paper we present \texttt{tstrait}, a Python library that
efficiently simulates quantitative traits on an arbitrary ARG.
% in \texttt{tskit} format.
\texttt{Tstrait} can
quickly simulate quantitative traits for population-scale datasets,
with a very low memory overhead, and taking into account
the rich historical detail contained within an ARG.
The \texttt{tstrait} library also integrates well with the wider
Python data-science ecosystem~\citep{numpy},
allowing users to efficiently analyse
large-scale data using familiar and ergonomic tools.
